---
title: "ODE Example with Springs"
author: "Graeme Keleher"
date: "May 29, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Intro
This is a simple example using the deSolve package to model the motion of springs. The following webpage has an excellent walkthrough of the thoery

http://tutorial.math.lamar.edu/Classes/DE/Vibrations.aspx


#Load the deSolve Package 
As well as ggplot2 for graphing
```{r message=FALSE, warning=FALSE}
library(deSolve)
library(ggplot2)
```


#Free Undamped Vibration
The simplest senario
```{r}

#Define the model parameters with k a the spring constant and m the mass
parameters = c(k = 1,m = 1)

#Define Model
Spring <- function(t, States, parameters) {
  with(as.list(c(States, parameters)),
       list(c(v,
              -(k/m)*y))
  )
}

#start with spring in position 0 and -6 velocity
States = c(y = 0, v = -6)

#Define a vector specifying the time intervals
times <- seq(0,40, by = 0.005)

#Run Model
result <- ode(y = States, times = times, func = Spring, parms = parameters)
```


```{r, echo = FALSE}
result.df = as.data.frame(result)

ggplot(result.df, aes(time, y )) + geom_line() + theme_bw() + labs(title = "Free, Undamped Vibrations", x = "Time (in seconds)", y = "Distance (in cm)")
```


#Free, Damped Vibrations
We now add a damping constant g. 
```{r}

parameters = c(k = 1,m = 1, g = .2)

Spring <- function(t, States, parameters) {
  with(as.list(c(States, parameters)),
       list(c(v,
              -(k/m)*y) - (g/m)*v)
  )
}

#start with spring in position 5 and zero velocity
States = c(y = 5, v = 0)

times <- seq(0,40, by = 0.005)

damped <- ode(y = States, times = times, func = Spring, parms = parameters)
```



```{r, echo = FALSE}
damped.df = as.data.frame(damped)

ggplot(damped.df, aes(time, y )) + geom_line() + theme_bw() + labs(title = "Free, Damped Vibrations", x = "Time (in seconds)", y = "Distance (in cm)")
```

#Undamped, Forced Vibrations
We add a forcing term which is a cosine function.
```{r}
parameters = c(k = 1,m = 1, g = 0, w = 1)

Spring <- function(t, States, parameters) {
  with(as.list(c(States, parameters)),
       list(c(v,
              -(k/m)*y) - (g/m)*v + cos(t*w))
  )
}

#start with spring in position 5 and zero velocity
States = c(y = 5, v = 0)

times <- seq(0,40, by = 0.005)

forced <- ode(y = States, times = times, func = Spring, parms = parameters)
```



```{r, echo = FALSE}
forced.df = as.data.frame(forced)

ggplot(forced.df, aes(time, y )) + geom_line() + theme_bw() + labs(title = "Forced, Undamped Vibrations", x = "Time (in seconds)", y = "Distance (in cm)")
```
#Damped, Forced Vibrations
We add a forcing term which is a cosine function.
```{r}
parameters = c(k = 1,m = 1, g = .1, w = .5)

Spring <- function(t, States, parameters) {
  with(as.list(c(States, parameters)),
       list(c(v,
              -(k/m)*y) - (g/m)*v + cos(t*w))
  )
}

#start with spring in position 5 and zero velocity
States = c(y = 5, v = 0)

times <- seq(0,100, by = 0.005)

forced_damp <- ode(y = States, times = times, func = Spring, parms = parameters)
```



```{r, echo = FALSE}
forced_damp.df = as.data.frame(forced_damp)

ggplot(forced_damp.df, aes(time, y )) + geom_line() + theme_bw() + labs(title = "Forced, Damped Vibrations", x = "Time (in seconds)", y = "Distance (in cm)")
```

